---
title: "Advanced Usage: reticulate-backed TDT Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Usage: reticulate-backed TDT Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tdtr)
```

This vignette is for workflows where the getting-started interface is not
enough: large recordings, direct Python `tdt` reader controls, live Python
objects, event-aligned reads, and memory diagnostics.

```{r}
example_path <- tdtr_example_block_path()
```

## Python State And Saving

Since TDT data are read using Python under the hood, there are some details to
consider when saving analysis state. `read_block()` returns a `tdt_block_py`
object by default. This object is a small R handle to a live Python `tdt` object
in the current R session. In this context, "handle" means the R object points to
data managed by Python. It does not mean all of the underlying stream arrays are
stored inside the R object.

That design is useful because it avoids copying every stream into R when you
only want to inspect names, metadata, or a bounded subset. It also means ordinary
R save functions should not be used as the durable record of a Python-backed
block. Saving the handle does not save the associated Python session.

A durable workflow has two parts:

1. save the path and read parameters needed to recreate the Python-backed block;
2. save explicitly collected R objects when you need durable analysis data.

```{r}
read_plan <- list(
  block_path = example_path,
  args = list(
    evtype = c("epocs", "streams"),
    t1 = 0,
    t2 = 5,
    verbose = 0
  )
)

block <- do.call(
  read_block,
  c(list(block_path = read_plan$block_path), read_plan$args)
)

class(block)
stream_names(block)
epoc_names(block)
summary(block)$streams
```

Save a read plan or a collected subset, not the live Python-backed handle as
long-term state.

```{r, eval = FALSE}
saveRDS(read_plan, "tdt-read-plan.rds")

selected <- collect_block(block, stores = c("_405A", "_465A"))
saveRDS(selected, "selected-tdt-block.rds")
```

## Python-side Processing

For Python-backed objects, `stream()` returns a live Python stream object. The
stream's `data` field is a NumPy array. An advanced R/Python user can do work on
that array in Python and return only a small result to R.

```{r}
py_stream <- stream(block, "_465A")
py_data <- py_stream$data

class(py_data)
```

This example computes small quality-control summaries in Python. The full trace
stays on the Python side; only the small dictionary of results crosses back into
R.

```{r}
reticulate::py_run_string("
import numpy as np

def qc_trace(x):
    x = np.asarray(x)
    return {
        'n_samples': int(x.shape[0]),
        'mean': float(np.nanmean(x)),
        'sd': float(np.nanstd(x)),
        'p01': float(np.nanpercentile(x, 1)),
        'p99': float(np.nanpercentile(x, 99)),
    }
")

reticulate::py$qc_trace(py_data)
```

This pattern is useful when the next few operations are naturally NumPy/Python
operations, or when returning a small result to R avoids copying a large array.

## Reader Controls

`read_block()` exposes common `tdt.read_block()` controls directly. This
example uses `evtype`, `channel`, `t1`, and `t2`; the event-aligned example
below uses `ranges`. Other controls, including `store`, `headers`, `nodata`,
`sortname`, and export options, are documented in `?read_block_py` and are
forwarded to Python `tdt` rather than reimplemented in R.

```{r}
bounded <- read_block(
  example_path,
  evtype = "streams",
  channel = 1,
  t1 = 1,
  t2 = 3,
  verbose = 0
)

stream_names(bounded)
summary(bounded)$streams
```

Extra keyword arguments can be passed through `...` when the underlying Python
package supports an option that the R wrapper has not named explicitly.

## Store Filters And Names

Read-time store filtering is delegated to Python `tdt`. This means store-name
behavior can expose details of the underlying Python package. In this packaged
fixture, filtering to `Fi1r` works as expected.

```{r}
fi1r_only <- read_block(
  example_path,
  evtype = "streams",
  store = "Fi1r",
  t1 = 1,
  t2 = 3,
  verbose = 0
)

stream_names(fi1r_only)
summary(fi1r_only)$streams
```

The same read-time filter does not currently select the leading-underscore
photometry stores by the stream names returned after a general read.

```{r}
photometry_filter <- read_block(
  example_path,
  evtype = "streams",
  store = "_465A",
  t1 = 1,
  t2 = 3,
  verbose = 0
)

stream_names(photometry_filter)
```

That behavior should be visible rather than hidden. Until the wrapper provides a
higher-level normalization for this case, use a bounded read and then select the
stream by name at collection time.

```{r}
photometry <- collect_stream(
  bounded,
  "_465A",
  as = "list",
  include_time = TRUE
)

list(
  samples = nrow(photometry$data),
  channels = ncol(photometry$data),
  first_times = head(photometry$time, 3)
)
```

## Explicit Collection

Collection is the point where data become ordinary R objects.

```{r}
signal <- collect_stream(bounded, "_465A", as = "list", include_time = TRUE)

list(
  samples = nrow(signal$data),
  channels = ncol(signal$data),
  first_times = head(signal$time, 3),
  r_size = format(utils::object.size(signal$data), units = "auto")
)
```

Use `collect_block()` when you want a durable R object for a selected subset.
The collected object no longer depends on a live Python session.

```{r}
selected <- collect_block(bounded, stores = "_465A")

class(selected)
summary(selected)$streams
```

## Event-aligned Reads

Event tables can be translated to the `2 x N` range matrix expected by Python
`tdt`. This lets you define event windows in R and still reduce the read before
collecting stream samples.

```{r}
ticks <- collect_epocs(block, store = "Tick")
ranges <- ranges_from_epocs(head(ticks, 2), pre = -0.1, post = 0.2)
ranges
```

```{r}
aligned <- read_block(
  example_path,
  evtype = "streams",
  ranges = ranges,
  verbose = 0
)

summary(aligned)$streams
```

## Profiling Before Scaling Up

Use `profile_tdt_memory()` on a bounded read before trying to load a large
recording. It profiles the read, metadata accessors, optional summary, optional
stream collection, and event collection.

```{r}
profile_tdt_memory(
  example_path,
  evtype = c("epocs", "streams"),
  t1 = 0,
  t2 = 5,
  stream = "_465A",
  verbose = 0
)
```

Interpret the result as a local diagnostic, not as exact total process memory.
The helper uses `Rprofmem()` for R allocations and Python `tracemalloc` for
Python allocations. Those tools do not perfectly measure every allocation made
by R, reticulate, NumPy, Python C extensions, or the operating system. Treat the
output as a way to compare alternative reads on the same machine and in the same
session. Differences smaller than roughly 10-30% should be treated cautiously;
large differences are usually informative.

The expected shape is:

- `metadata_accessors` stays small because it reads names and metadata;
- `collect_stream:*` allocates R memory because it intentionally copies selected
  stream samples into R;
- the Python peak column is useful context, but it is not the same as total
  resident memory for the R process.

Use the high-level workflow when selected data fit comfortably in memory and the
downstream analysis expects matrices or tibbles. Use the Python-backed workflow
when you need to inspect first, reduce at read time, keep large arrays on the
Python side, or persist a reproducible read plan separately from collected R
analysis data.
